Symmetric Peer-to-Peer Applications

### 2022-10-01 tpd-22.tex
...

\usepackage{xspace}
\newcommand{\FC}       {Freechains\xspace}
\newcommand{\reps}     {\emph{reps}\xspace}
\newcommand{\onerep}   {\emph{1~rep}\xspace}
\newcommand{\nreps}[1] {\emph{#1~reps\xspace}}
\newcommand{\code}[1]  {\texttt{\footnotesize{#1}}}
\newcommand{\Xon} {$1{\rightarrow}N$\xspace}
\newcommand{\Xno} {$1{\leftarrow}N$\xspace}
\newcommand{\Xnn} {$N{\leftrightarrow}N$\xspace}
\newcommand{\Xoo} {$1{\leftrightarrow}1$\xspace}
\newcommand{\Xo}  {$1{\hookleftarrow}$\xspace}
\newcommand{\lin}[1]{(\emph{ln. #1}\xspace)}

@@

In Figure~\ref{fig.middleware}, a peer is depicted as an application that uses
an API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
the other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp in the future, such that all peers can
satisfy.
The middleware controls the execution of the application by issuing ticks and
synchronized events.

@@

\subsubsection{Events Broadcasting}
\label{sec.tml.middleware.events}

As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
peers with a timestamp in the future such that all peers are able to apply them
in sync.
%
To prevent broadcast cycles, event are tagged with a source peer and sequence
number.
Each peer keeps a vector with the highest sequence number it received from each
other peer, such that lower numbers can be ignored when received.
Higher numbers update the vector and trigger a broadcast to neighbours.
Since we rely on TCP connections, it is not possible to receive out-of-order
sequence numbers from the same source peer.
%
As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
themselves unique identifiers in a contiguous range.
The reasons are
    (a) to fit a \code{uint8\_t} size, and
    (b) to use a contiguous vector of sequence numbers.
Therefore, this is not a serious limitation, and a separate mechanism could be
used for assignment, with few modifications to core of the middleware.

@@

{\footnotesize
\begin{verbatim}
typedef struct {
    uint8_t  src;       // source peer
    uint32_t seq;       // sequence number
    uint32_t tick;      // tick to apply
    p2p_evt  evt;       // event id + payload
} p2p_pak;

@@

{\footnotesize
\begin{verbatim}
01  void p2p_loop (...,cb_ini,cb_sim,cb_ren,cb_evt) {
02    cb_ini(1);
03    cb_sim(<P2P_EVT_START>);
04
05    while (<app-running>) {
06      if (<next-packet>) {
07        // fast forward if I'm late
08        if (<packet-sync-future>) {
09          p2p_travel(<now>, <packet-sync-time>);
10        }
11
12        // travel back & forth if I'm early
13        if (<packet-evt-past>) {
14          p2p_travel(<now>, <packet-evt-time>);
15          p2p_travel(<packet-evt-time>, <now>);
16        }
17
18        // simulate in real time if I'm on time
19        if (<packet-evt-ok>) {
20          cb_sim(<packet-evt>);
21          cb_ren();
22        }
23      }
24
25      if (<next-input>) {
26        // broadcast event from local
27        p2p_evt evt;
28        if (cb_evt(&evt)) {
29          p2p_bcast(<future>, &evt);
30        }
31      }
32
33      // simulate tick in real time
34      if (<next-tick>) {
35        cb_sim(<P2P_EVT_TICK>);
36        <application-snapshot-every-second>
37        cb_ren();
38      }
39    }
40
41    cb_ini(0);
42  }
\end{verbatim}
}

@@

Regarding network packets, if we receive a time synchronization packet
\code{SYNC} from the future \lin{7--10}, then the we fast forward the
application to catch up in time (Section~\ref{sec.tml}.b).
If we receive an event that should have been applied in the past \lin{12--16},
then we travel back and forth (Section~\ref{sec.tml}.a).
Otherwise, we just apply the event in real time \lin{18--22}
(Section~\ref{sec.tml.api.cb_sim}).
We detail the function \code{p2p\_travel} in
Section~\ref{sec.tml.middleware.travel}.
%
Regarding local inputs, we call \code{cb\_evt} \lin{26--30} to signal if the
event should be broadcast, also transforming it to an application event
(Section~\ref{sec.tml.api.cb_evt}).
%
Regarding time tick, we call \code{cb\_sim} and \code{cb\_ren} in real time
\lin{33--38}.
The middleware also takes periodic snapshots of the application state to
rollback and assist time travelling (Section~\ref{sec.tml.api.state}).

@@

{\footnotesize
\begin{verbatim}
01  void p2p_travel (int from, int to) {
02    for (i=from..to) {
03      int bef = <tick-of-snapshot-before-i>
04      <restore-snapshot-at-bef>
05      for (j=bef..i) {
06        cb_sim(<tick-or-evt-at-j>)
07      }
08      cb_ren();
09      <delay-ms>
10    }
11  }
\end{verbatim}
}

The function \code{p2p\_travel} time travels, starting at tick \code{from},
tick by tick, until it reaches tick \code{to} \lin{2--10}.
The loop can go in any direction in time, i.e., \code{from} may be higher than
\code{to}.
At each step, we need to find the tick \code{bef} with the closest past
snapshot \lin{3}, restore it \lin{4}, and than simulate the remaining steps
from the snapshot to the desired tick \lin{5--7}.
We also exhibit each step with a small delay \lin{8--9} so that users
experience smooth time transitions.

@@

This section provided an overview of the middleware implementation to answer,
in concrete terms, how it ensures the desired properties enumerated in the
Introduction and at the beginning of Section~\ref{sec.tml.middleware}:
    (a) peers meet event deadlines either in real time \lin{18--21} or
        travelling back and forth in time \lin{12--16}
        (Section~\ref{sec.tml.middleware.loop});
    (b) peers synchronize their clocks with periodic \code{SYNC} packets that
        fast forwards late peers \lin{7--10}
        (Section~\ref{sec.tml.middleware.loop}); and
    (c) peers can join the network at any time and still catch up in time by
        receiving the full event history
        (Section~\ref{sec.tml.middleware.events}).

@@

In this section, we evaluate if (and to what extent) our design meets its goals
to ensure that all peers (a) meet event deadlines, (b) advance in sync in real
time, and (c) can leave and join the network and remain symmetric.
%
We simulate an application varying the network topology, latency, and churn, as
well as the frequency and deadlines of events.
%
We then measure (a) the recurrence of time travels, (b) the real-time pace of
peers, and (c) the recoverability from churns, such that they validate our
goals above, respectively.

Regarding (a), if a peer experiences a time travel, it means that it received
an event after its deadline.
We count all time travel occurrences during the simulation.
%
Regarding (b), we use an absolute simulation clock to track, in real time, the
execution pace of each peer in comparison to the reference peer, which is the
one most advanced in time.
%
Regarding (c), we make peers leave and join the network periodically, possibly
creating partitions, and assert that they all catch up in real time.

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{topo}
  \caption{
    \label{fig.topo}
    Network topology: 21 peers organized as a straight line ($0--3$), a simple
    cycle ($3--8$), a second straight line ($8--11$), a full connected cycle
    ($12--17$), and a third straight line ($17--20$).
    An optional link ($0,20$) creates a long cycle in the network.
  }
\end{figure*}

@@

We now detail how we instrument the middleware implementation to measure the
properties above:
%
\begin{enumerate}
%
\item \textbf{Recurrence of time travels:}
Every time a peer is ahead of time and needs to travel back, we output its id
and how much ticks it needs to travel back.
At the end, we measure the percentage of the sum of all time travel ticks over
the total simulation ticks (\code{TODO} in Figure~\ref{TODO}).
As an example, if all peers sum $1000$ time travel ticks over $100k$
accumulated simulation ticks, then the final measure is $1\%$.
%
\item \textbf{Real-time pace of peers:}
Every time a peer receives \code{SYNC} packets and is behind, we output its id
and how much ticks it needs to fast forward.
At the end, we measure the percentage of the sum of all forward ticks over the
total simulation ticks (\code{TODO} in Figure~\ref{TODO}).
As an example, if all peers sum $1000$ forward ticks over $100k$ accumulated
simulation ticks, then the final measure is $1\%$.
%
We also start each peer with a random delay of up to $10s$ to force a
synchronization mismatch from the beginning.
%
Each peer outputs its current tick continuously, allowing us to compare them
in the shared timeline.
Each time we see a greater tick never seen in the timeline before, we count
smaller ticks from other peers (considering latency), which would constitute
time violations (\code{TODO} in Figure~\ref{TODO}).
We expect this counter to be close to zero even considering the whole
simulation.
%
\item \textbf{Correctness of unstable peers:}
We run a separate simulation in which, for every $100s$, each peer remains
$40s$ online and $20s$ offline, including peers \code{9--11} in the middle of
the network.
This creates partitions in which sides advance in time and generate events out
of sync, which need to be merged and resynchronized constantly.
We measure
    - state (even for stable peers)
    - qto tempo leva para sync
%
\end{enumerate}

>>>>

\usepackage{xspace}
\newcommand{\FC}       {Freechains\xspace}
\newcommand{\reps}     {\emph{reps}\xspace}
\newcommand{\onerep}   {\emph{1~rep}\xspace}
\newcommand{\nreps}[1] {\emph{#1~reps\xspace}}
\newcommand{\code}[1]  {\texttt{\footnotesize{#1}}}
\newcommand{\Xon} {$1{\rightarrow}N$\xspace}
\newcommand{\Xno} {$1{\leftarrow}N$\xspace}
\newcommand{\Xnn} {$N{\leftrightarrow}N$\xspace}
\newcommand{\Xoo} {$1{\leftrightarrow}1$\xspace}
\newcommand{\Xo}  {$1{\hookleftarrow}$\xspace}
\newcommand{\lin}[1]{(\emph{ln. #1}\xspace)}
\newcommand{\linx}[1]{\emph{ln. #1}\xspace}

@@

In Figure~\ref{fig.middleware}, a peer is depicted as an application that uses
an API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
the other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp scheduled to the future, such that all
peers can satisfy.
The middleware controls the execution of the application by issuing ticks and
synchronized events.

@@

\subsubsection{Event Broadcasting}
\label{sec.tml.middleware.events}

As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
peers with a timestamp scheduled to the future such that all peers are able to
apply them in sync.
%
To prevent broadcast cycles, each peer keeps a vector with the highest
timestamp it received from each other peer, such that lower numbers can be
ignored when received.
Since we rely on TCP connections, it is not possible to receive out-of-order
timestamps attached to a given source peer.
When the middleware receives a higher number from a given peer, it updates its
vector position and triggers a broadcast to neighbours.
%
As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
themselves unique identifiers in a contiguous range, such that they fit in a
contiguous vector of timestamps.
Therefore, this is not a severe limitation, and a separate mechanism could be
used for assignments, with few modifications to core of the middleware.

@@

{\footnotesize
\begin{verbatim}
typedef struct {
    uint8_t  src;       // source peer
    uint32_t tick;      // tick to apply
    p2p_evt  evt;       // event id + payload
} p2p_pak;

@@

{\footnotesize
\begin{verbatim}
01  void p2p_loop (...,cb_ini,cb_sim,cb_ren,cb_evt) {
02    cb_ini(1);    // initialization
03    cb_sim(<P2P_EVT_START>);
04
05    while (<app-running>) {
06      if (<next-network-packet>) {
07        // fast forward if I'm late
08        if (<packet-sync-future>) {
09          p2p_travel(<now>, <packet-sync-time>, <speed>);
10        }
11
12        // travel back & forth if I'm early
13        if (<packet-evt-past>) {
14          p2p_travel(<now>, <packet-evt-time>, <speed>);
15          p2p_travel(<packet-evt-time>, <now>, <speed>);
16        }
17
18        // real time if I'm on time
19        if (<packet-evt-ok>) {
20          cb_sim(<packet-evt>);
21          cb_ren();
22        }
23      }
24
25      if (<next-local-input>) {
26        // broadcast local event
27        p2p_evt evt;
28        if (cb_evt(&evt)) {
29          p2p_bcast(<future>, &evt);
30        }
31      }
32
33      // simulate tick in real time
34      if (<next-local-tick>) {
35        cb_sim(<P2P_EVT_TICK>);
36        <application-snapshot-every-second>
37        cb_ren();
38      }
39    }
40
41    cb_ini(0);    // finalization
42  }
\end{verbatim}
}

@@

Regarding network packets, if we receive a time synchronization packet
\code{SYNC} from the future \lin{7--10}, then we fast forward the application
to catch up in time (Section~\ref{sec.tml}.b).
If we receive an event that should have been applied in the past \lin{12--16},
then we travel back and forth (Section~\ref{sec.tml}.a).
Otherwise, we just apply the event in real time \lin{18--22}
(Section~\ref{sec.tml.api.cb_sim}).
We detail the function \code{p2p\_travel} in
Section~\ref{sec.tml.middleware.travel}.
%
Regarding local inputs, we call \code{cb\_evt} \lin{26--30} to signal if the
event should be broadcast, also transforming it to an application event
(Section~\ref{sec.tml.api.cb_evt}).
%
Regarding time ticks, we call \code{cb\_sim} and \code{cb\_ren} in real time
\lin{33--38}.
The middleware also takes periodic snapshots of the application state to
optimize time travelling (Section~\ref{sec.tml.api.state}).

@@

{\footnotesize
\begin{verbatim}
01  void p2p_travel (int from, int to, int ms) {
02    for (i=from..to) {
03      int bef = <tick-of-snapshot-before-i>
04      <restore-snapshot-at-bef>
05      for (j=bef..i) {
06        cb_sim(<tick-or-evt-at-j>)
07      }
08      cb_ren();
09      <delay-ms>
10    }
11  }
\end{verbatim}
}

The function \code{p2p\_travel} time travels the application, starting at tick
\code{from}, tick by tick, until it reaches tick \code{to} \lin{2--10}.
The loop can travel in both directions, i.e., \code{from} may be higher than
\code{to}.
At each step, we need to find the tick \code{bef} with the closest past
snapshot \lin{3}, restore it \lin{4}, and than simulate the remaining steps
from the snapshot to the desired tick \lin{5--7}.
We also exhibit each step with a small \code{ms} delay \lin{8--9} so that users
experience smooth time transitions.

@@

This section provided an overview of the middleware implementation to answer,
in concrete terms, how it ensures the desired properties enumerated in the
Introduction and at the beginning of Section~\ref{sec.tml.middleware}:
    (a) peers meet event deadlines either in real time \lin{18--21} or
        travelling back and forth in time
        (\linx{12--16} / \ref{sec.tml.middleware.loop});
    (b) peers synchronize their clocks with periodic \code{SYNC} packets that
        fast forwards late peers
        (\linx{7--10} / \ref{sec.tml.middleware.loop}); and
    (c) peers can join the network at any time and still catch up in time by
        receiving the full event history
        (Section~\ref{sec.tml.middleware.events}).

@@

In this section, we evaluate if (and to what extent) our design meets its goals
to ensure that all peers (a) meet event deadlines, (b) advance in sync in real
time, and (c) can leave and join the network and remain symmetric.
%
We simulate an application varying the network latency and churn, as well as
the frequency and deadlines of events.
%
We then measure (a) the recurrence of time travels, (b) the real-time pace of
peers, and (c) the recoverability from churns, such that they validate our
goals above, respectively.

Regarding (a), if a peer experiences a time travel, it means that it received
an event after its deadline.
We then count all time travel occurrences during the simulation.
%
Regarding (b), we use an absolute simulation clock to track, in real time, the
execution pace of each peer.
% in comparison to the reference peer, which is the one most advanced in time.
We then count the mismatches between the clocks.
%
Regarding (c), we make peers leave and join the network periodically, possibly
creating partitions.
We then assert that they all catch up in real time.

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{topo}
  \caption{
    \label{fig.topo}
    Network topology: 21 peers organized as a straight line ($0-3$), a simple
    cycle ($3-8$), a second straight line ($8-11$), a full connected cycle
    ($12-17$), and a third straight line ($17-20$).
    %An optional link ($0,20$) creates a long cycle in the network.
  }
\end{figure*}

@@

We now detail how we instrument the middleware implementation to measure the
properties above:
%
\begin{enumerate}
%
\item \textbf{Recurrence of time travels:}
Every time a peer is ahead of time and needs to travel back, we output its id
and how much ticks it needs to travel back.
At the end, we measure the percentage of the sum of all time travel ticks over
the total simulation ticks (\code{TODO} in Figure~\ref{TODO}).
As an example, if all peers sum $1000$ time travel ticks over $100k$
accumulated simulation ticks, then the final measure is $1\%$.
TODO: in the figure, we consider range a-b, c-d...
%
\item \textbf{Real-time pace of peers:}
Every time a peer receives \code{SYNC} packets and is behind, we output its id
and how much ticks it needs to fast forward.
At the end, we measure the percentage of the sum of all forward ticks over the
total simulation ticks (\code{TODO} in Figure~\ref{TODO}).
As an example, if all peers sum $1000$ forward ticks over $100k$ accumulated
simulation ticks, then the final measure is $1\%$.
%
We also start each peer with a random delay of up to $10s$ to force a
synchronization mismatch from the beginning.
%
Each peer outputs its current tick continuously, allowing us to compare them
in the shared timeline.
Each time we see a greater tick never seen in the timeline before, we count
smaller ticks from other peers (considering latency), which would constitute
time violations (\code{TODO} in Figure~\ref{TODO}).
We expect this counter to be close to zero even considering the whole
simulation.
TODO: in the figure, we consider range a-b, c-d...
TODO: in comparison to the reference peer, which is the one most advanced in time.
%
\item \textbf{Correctness of unstable peers:}
We run a separate simulation in which, for every $100s$, each peer remains
$40s$ online and $20s$ offline, including peers \code{9--11} in the middle of
the network.
This creates partitions in which sides advance in time and generate events out
of sync, which need to be merged and resynchronized constantly.
We measure
    - state (even for stable peers)
    - qto tempo leva para sync
TODO: in the figure, we consider range a-b, c-d...
%
\end{enumerate}
